---
title: "Chapter 1: Statistical Foundations and Data Concepts"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: flatly
    highlight: tango
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  comment = "#>"
)
```

------------------------------------------------------------------------

# Populations, Samples, and Parameters

## Population vs Sample

A **population** is the entire group you want to study.\
A **sample** is a smaller, manageable subset drawn from that population for analysis.

> **Why sample?** Studying an entire population is often impractical or impossible. A well-chosen sample lets us draw conclusions about the population without measuring every member.

## Parameters vs Statistics

| Term | Definition | Example |
|----|----|----|
| **Parameter** | A numerical value that describes the *population* | Population mean μ |
| **Statistic** | A numerical value calculated from a *sample* | Sample mean x̄ |

Parameters are typically unknown; statistics are what we compute and use to *estimate* parameters.

## Sampling Variability and Randomness

**Sampling variability** is the natural difference you see between statistics from different samples drawn from the same population. No two samples are exactly alike.

**Randomness** ensures that every member of the population has a fair, unbiased chance of being selected, which is what makes our sample representative.

------------------------------------------------------------------------

# Data Types and Structure

## Categorical vs Numerical Data

-   **Categorical data** — represents groups or labels (e.g., blood type, gender, country).
-   **Numerical data** — represents measurable quantities (e.g., weight, temperature, salary).

## Discrete vs Continuous Variables

| Variable Type | Description | Examples |
|----|----|----|
| **Discrete** | Countable whole numbers; gaps exist between values | Number of students, coin flips |
| **Continuous** | Any value within a range; decimals allowed | Height, temperature, reaction time |

## Measurement Scales

The scale of measurement determines what statistical operations are meaningful.

| Scale | Properties | Examples |
|----|----|----|
| **Nominal** | Categories only; no order | Eye colour, gender |
| **Ordinal** | Ordered categories; differences not meaningful | Satisfaction level, class rank |
| **Interval** | Equal gaps; no true zero | Temperature in °C, calendar year |
| **Ratio** | Equal gaps + true zero | Weight, height, age |

> **Tip:** The higher the scale, the more statistical techniques you can apply. Ratio data supports the widest range of analyses.

------------------------------------------------------------------------

# R Foundations

## Creating Data and Sampling with `sample()`

The most fundamental structure in R is the **vector** — an ordered collection of values of the same type.

```{r vectors-and-sampling}
# Create a simple numeric vector using c()
x <- c(5, 10, 15, 20)
x

# Draw 10 random numbers from 1–100 (with replacement)
# replace = TRUE allows the same number to appear more than once
set.seed(42)                          # set seed for reproducibility
sample_data <- sample(1:100, size = 10, replace = TRUE)
sample_data
```

**Why `set.seed()`?** It fixes the random number generator so your results are reproducible — anyone running the same code gets the same output.

## Basic Summaries with `summary()`

`summary()` gives you a five-number summary (plus the mean) in one call — a great first step when exploring new data.

```{r summary-demo}
data <- c(12, 15, 20, 25, 30, 35, 40)

summary(data)
```

| Output  | Meaning                        |
|---------|--------------------------------|
| Min.    | Smallest value                 |
| 1st Qu. | 25th percentile                |
| Median  | Middle value (50th percentile) |
| Mean    | Arithmetic average             |
| 3rd Qu. | 75th percentile                |
| Max.    | Largest value                  |

## Initial Exploratory Plots

Visualising data before formal analysis helps you spot patterns, outliers, and the overall shape of the distribution.

```{r exploratory-plots, fig.align='center', fig.width=8, fig.height=4}
set.seed(42)
data <- sample(1:100, size = 50, replace = TRUE)

# --- Side-by-side plots ---
par(mfrow = c(1, 2))

# Histogram: shows the frequency/distribution of values
hist(data,
     main  = "Histogram of Sample Data",
     xlab  = "Value",
     col   = "steelblue",
     border = "white")

# Boxplot: highlights median, spread, and potential outliers
boxplot(data,
        main   = "Boxplot of Sample Data",
        ylab   = "Value",
        col    = "lightcoral",
        border = "darkred")

par(mfrow = c(1, 1))   # reset layout
```

### What to look for

-   **Histogram** — Is the distribution symmetric, skewed left, or skewed right? Any unusual gaps?\
-   **Boxplot** — Where is the median (centre line)? Are there points outside the whiskers (outliers)?

------------------------------------------------------------------------

# Chapter Summary

| Topic | Key Takeaway |
|----|----|
| Population vs Sample | Samples are used to make inferences about populations |
| Parameters vs Statistics | Parameters describe populations; statistics describe samples |
| Sampling Variability | Different samples from the same population yield different statistics |
| Data Types | Categorical (labels) vs Numerical (quantities) |
| Variable Types | Discrete (countable) vs Continuous (measurable) |
| Measurement Scales | Nominal → Ordinal → Interval → Ratio (increasing power) |
| R: `c()`, `sample()` | Build vectors and draw random samples |
| R: `summary()` | Quick five-number summary for any numeric vector |
| R: `hist()`, `boxplot()` | Core tools for initial visual exploration |
