---
title: "Statistical Computing — Week 8: KDE, Classification & Mixture Models"
subtitle: "Complete Lab Solutions with Exam-Ready Explanations"
author: "Week 8 Lab"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    highlight: tango
    latex_engine: xelatex
    extra_dependencies: ["amsmath", "amssymb", "float"]
header-includes:
  - \usepackage{framed}
  - \usepackage{xcolor}
  - \definecolor{examtip}{RGB}{0, 100, 150}
  - \definecolor{shadecolor}{RGB}{230, 240, 250}
  - \newcommand{\examtip}[1]{\begin{snugshade}\textcolor{examtip}{\textbf{Exam Tip:}} #1\end{snugshade}}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 7, fig.height = 5,
                      fig.pos = "H", out.width = "85%", fig.align = "center")
options(digits = 4)
```

\newpage

# Background Theory You Need to Know

Before diving into code, here is the theory behind every technique in this lab. **Understanding these ideas is what gets you marks in an exam.**

## What is Kernel Density Estimation (KDE)?

A histogram is a rough way to estimate the shape of data. KDE is a **smooth** version. The idea:

- Place a small "bump" (called a **kernel**) on top of every data point.
- Add all the bumps together and normalise.
- The result is a smooth curve that estimates the true probability density.

Mathematically, for $n$ data points $x_1, \ldots, x_n$:

$$\hat{f}(x) = \frac{1}{n h} \sum_{i=1}^{n} K\left(\frac{x - x_i}{h}\right)$$

where:

- $K$ is the **kernel function** (usually Gaussian: $K(u) = \frac{1}{\sqrt{2\pi}} e^{-u^2/2}$)
- $h$ is the **bandwidth** — this controls how smooth the estimate is

\examtip{The bandwidth $h$ is the single most important choice. Too small $\to$ wiggly/noisy. Too large $\to$ oversmoothed, hides real features. Bandwidth selection methods: ``nrd0'' (rule of thumb), ``ucv'' (cross-validation).}

## What is a Nonparametric Bayes Classifier?

Given two classes (e.g. smoker/non-smoker), we want $P(\text{class} = 1 \mid x)$. By Bayes' theorem:

$$P(Y = 1 \mid x) = \frac{\pi_1 \, \hat{f}_1(x)}{\pi_0 \, \hat{f}_0(x) + \pi_1 \, \hat{f}_1(x)}$$

where:

- $\hat{f}_0(x)$ = KDE of $x$ for class 0 observations
- $\hat{f}_1(x)$ = KDE of $x$ for class 1 observations  
- $\pi_1 = P(Y=1)$ = prior probability (proportion of class 1 in data)
- $\pi_0 = 1 - \pi_1$

We classify as class 1 if $P(Y=1 \mid x) > 0.5$.

\examtip{There are two ways to compute the denominator: (1) the \textbf{mixture} denominator $\pi_0 f_0 + \pi_1 f_1$, or (2) the \textbf{marginal} density $f(x)$ estimated from ALL data. Both should give similar results. The mixture version is theoretically exact.}

## What is a Mixture Model?

A **mixture model** says the data comes from $K$ different sub-populations (components), each with its own distribution. For a two-component Gaussian mixture:

$$f(x) = \lambda \, \phi(x; \mu_1, \sigma_1) + (1 - \lambda) \, \phi(x; \mu_2, \sigma_2)$$

where $\phi(x; \mu, \sigma)$ is the Normal density with mean $\mu$ and standard deviation $\sigma$, and $\lambda$ is the **mixing proportion** (probability a random observation comes from component 1).

## The EM Algorithm

We can't directly maximise the likelihood of a mixture (the log of a sum is hard). Instead we iterate:

1. **E-step (Expectation):** For each data point, compute the *responsibility* $\gamma_i$ — the probability that point $i$ belongs to component 1:
$$\gamma_i = \frac{\lambda \, \phi(x_i; \mu_1, \sigma_1)}{\lambda \, \phi(x_i; \mu_1, \sigma_1) + (1-\lambda) \, \phi(x_i; \mu_2, \sigma_2)}$$

2. **M-step (Maximisation):** Update the parameters using weighted averages:
$$\mu_1 = \frac{\sum \gamma_i x_i}{\sum \gamma_i}, \quad \sigma_1^2 = \frac{\sum \gamma_i (x_i - \mu_1)^2}{\sum \gamma_i}, \quad \lambda = \frac{\sum \gamma_i}{n}$$

3. **Repeat** until convergence (parameters stop changing).

\examtip{The EM algorithm is \textbf{guaranteed} to increase the likelihood at every step, but can get stuck in local maxima. That's why \textbf{initial values matter} — different starting points can give different answers. Always try several.}

## AIC and BIC for Model Selection

How many components do we need? More components always fit better, but we risk **overfitting**. AIC and BIC penalise complexity:

$$\text{AIC} = -2 \log L + 2k, \qquad \text{BIC} = -2 \log L + k \log n$$

where $k$ = number of parameters, $n$ = sample size. **Lower is better.**

\examtip{BIC penalises more than AIC (because $\log n > 2$ when $n > 7$). BIC tends to prefer simpler models. For a $K$-component Normal mixture, $k = 3K - 1$ (each component has $\mu$, $\sigma$, plus $K-1$ free mixing proportions).}

\newpage

# Setup: Install and Load Packages

```{r packages, eval=FALSE}
# Run this ONCE to install
install.packages(c("mixtools", "MASS", "CircNNTSR", "stats4"),
                 repos = "https://cloud.r-project.org")
```

```{r load-packages}
library(mixtools)     # normalmixEM for fitting mixture models
library(MASS)         # galaxies dataset
library(stats4)       # mle() function
```

\newpage

# Build a Nonparametric Classifier

## Smoking / Birth Weight Classifier

We predict smoking status (`smoke`) from birth weight (`bwt`). The idea: estimate the density of birth weight separately for smokers and non-smokers, then use Bayes' theorem.

```{r classifier-load}
# Load the smoking/birthweight data
# UPDATE THIS PATH to wherever you saved the CSV from Moodle:
smoking <- read.csv("smoking.csv")  # columns: bwt, smoke (0/1)

# If you don't have the CSV, you can use the birthwt dataset from MASS:
# library(MASS)
# smoking <- data.frame(bwt = birthwt$bwt, smoke = birthwt$smoke)

head(smoking)
table(smoking$smoke)
```

### Step 1: Estimate the Densities

```{r classifier-densities}
lo <- min(smoking$bwt) - 100
hi <- max(smoking$bwt) + 100

# f  = density of bwt for ALL observations (marginal density)
# f0 = density of bwt for non-smokers (smoke = 0)
# f1 = density of bwt for smokers (smoke = 1)

f  <- density(smoking$bwt, from = lo, to = hi)
f0 <- density(smoking$bwt[smoking$smoke == 0], from = lo, to = hi)
f1 <- density(smoking$bwt[smoking$smoke == 1], from = lo, to = hi)
```

### Step 2: Plot the Densities

```{r classifier-density-plots, fig.height=7}
par(mfrow = c(2, 2), mar = c(4, 3, 2, 1) + 0.1, cex = 0.8)
plot(f,  main = "Density of bwt (all)",       lwd = 2)
plot(f0, main = "Density of bwt (smoke=0)", col = 2, lwd = 2)
plot(f1, main = "Density of bwt (smoke=1)", col = 3, lwd = 2)
plot(f0, main = "Both densities overlaid",  col = 2, lwd = 2)
lines(f1, col = 3, lwd = 2)
legend("topright", legend = c("Non-smoker", "Smoker"),
       col = c(2, 3), lwd = 2, cex = 0.8)
```

**What to look for:** If the two densities are well-separated, the classifier will work well. If they overlap heavily, classification will be poor (which is expected here — birth weight alone is a weak predictor of smoking).

### Step 3: Compute Posterior Probability

Using Bayes' theorem:

$$P(\text{smoke}=1 \mid \text{bwt}) = \frac{\pi_1 \hat{f}_1(\text{bwt})}{\pi_0 \hat{f}_0(\text{bwt}) + \pi_1 \hat{f}_1(\text{bwt})}$$

```{r classifier-posterior}
p <- mean(smoking$smoke)  # prior P(smoke = 1)

par(mfrow = c(1, 2), mar = c(4, 4, 2, 1) + 0.1, cex = 0.8)

# Method 1: Mixture denominator (theoretically exact)
posterior_mix <- f1$y * p / (p * f1$y + (1 - p) * f0$y)
plot(f$x, posterior_mix, type = "l", lwd = 2,
     xlab = "Birth Weight (g)", ylab = "P(smoke=1 | bwt)",
     main = "Mixture Denominator")
abline(h = 0.5, lty = 3, col = "red")

# Method 2: Marginal denominator
posterior_marg <- f1$y * p / f$y
plot(f$x, posterior_marg, type = "l", lwd = 2,
     xlab = "Birth Weight (g)", ylab = "P(smoke=1 | bwt)",
     main = "Marginal Denominator")
abline(h = 0.5, lty = 3, col = "red")
```

**Interpretation:** The red dashed line at 0.5 is the classification boundary. Where the curve is above 0.5, we classify as "smoker". Both methods should give very similar results.

\examtip{If the posterior is mostly flat near the prior $\pi_1$, the predictor ($x$) carries very little information about the class. The classifier is barely better than guessing.}

\newpage

# Old Faithful: Multivariate Density Estimation

## Sequential Waiting Times

The idea: is the waiting time for the **next** eruption related to the **current** waiting time?

```{r faithful-data}
Waiting <- faithful$waiting

# x = current waiting time (drop the last one)
# y = NEXT waiting time (drop the first one)
x <- Waiting[-length(Waiting)]
y <- Waiting[-1]
```

### Scatterplot

```{r faithful-scatter}
plot(x, y, pch = 19, col = rgb(0, 0, 0.6, 0.5), cex = 0.8,
     xlab = "Current Waiting Time (min)",
     ylab = "Next Waiting Time (min)",
     main = "Sequential Waiting Times — Old Faithful")
```

**Comment:** You should see **two clusters** — one where both current and next waiting times are short, and one where both are long. There may also be "transition" points (short $\to$ long and long $\to$ short). This suggests the eruptions alternate between two regimes.

### 2D Kernel Density Estimate

```{r faithful-2d, fig.height=6}
# Compute 2D density
kde2d_fit <- kde2d(x, y, n = 50)

# Filled contour plot
par(mfrow = c(1, 2))
contour(kde2d_fit, nlevels = 15, lwd = 1.5,
        xlab = "Current Wait", ylab = "Next Wait",
        main = "2D KDE — Contour")

# Perspective (3D surface) plot
persp(kde2d_fit, theta = 30, phi = 25, shade = 0.4,
      xlab = "Current", ylab = "Next", zlab = "Density",
      main = "2D KDE — Surface", col = "lightblue")
```

**Comment:** The 2D density clearly shows **two modes** (peaks). This confirms the bimodal, clustered structure: Old Faithful alternates between two eruption regimes with distinct waiting times.

\examtip{2D KDE is done with \texttt{kde2d()} from the MASS package. It returns a grid of density values. Use \texttt{contour()}, \texttt{filled.contour()}, or \texttt{persp()} to visualise.}

\newpage

# Bridging Exercise: Galaxy Velocities

## Load and Explore

```{r galaxies-data}
library(MASS)
data(galaxies)

# galaxies is just a numeric vector of 82 velocities (km/s)
cat("Number of galaxies:", length(galaxies), "\n")
summary(galaxies)
```

## Histogram

```{r galaxies-hist}
hist(galaxies, breaks = 30, col = "grey85", border = "white",
     main = "Galaxy Velocities", xlab = "Velocity (km/s)",
     probability = TRUE)
```

**How many modes?** You should see roughly **3 to 4 modes** — a big cluster around 20,000, something near 10,000, and bumps around 23,000 and 34,000.

## Overlay Different KDE Estimates

```{r galaxies-kde, fig.height=6}
hist(galaxies, breaks = 30, col = "grey85", border = "white",
     probability = TRUE, ylim = c(0, 0.00018),
     main = "Galaxy Velocities — Comparing Bandwidths",
     xlab = "Velocity (km/s)")

# 1. Rectangular kernel, bandwidth = 500
d_rect <- density(galaxies, bw = 500, kernel = "rectangular")
lines(d_rect, col = "red", lwd = 2)

# 2. Gaussian kernel, rule-of-thumb bandwidth (nrd0)
d_nrd0 <- density(galaxies, bw = "nrd0")
lines(d_nrd0, col = "blue", lwd = 2)

# 3. Gaussian kernel, cross-validation bandwidth (ucv)
d_ucv <- density(galaxies, bw = "ucv")
lines(d_ucv, col = "darkgreen", lwd = 2)

legend("topright",
       legend = c(paste0("Rectangular, bw=500 (h=", round(500, 0), ")"),
                  paste0("Gaussian nrd0 (h=", round(d_nrd0$bw, 0), ")"),
                  paste0("Gaussian ucv (h=", round(d_ucv$bw, 0), ")")),
       col = c("red", "blue", "darkgreen"), lwd = 2, cex = 0.8)
```

**Discussion:**

- **Rectangular, bw=500:** Shows the most detail / modes, but can be noisy.
- **Gaussian nrd0:** The rule-of-thumb assumes unimodality, so it **oversmooths** multimodal data. Many bumps get hidden.
- **Gaussian ucv:** Cross-validation adapts to the data. It usually gives the clearest picture of the true multimodal structure.

\examtip{For multimodal data, the rule-of-thumb bandwidth (``nrd0'') is often too wide because it assumes a single Gaussian. Cross-validation (``ucv'' or ``SJ'') is better for discovering multiple modes.}

## How Many Mixture Components?

Based on the KDE, we can count roughly **3 modes** (some argue 4). So a 3-component Normal mixture is a reasonable starting point. Let's test this formally.

## Fit 2- and 3-Component Mixtures

```{r galaxies-mix}
set.seed(42)

# 2-component mixture
mix2 <- normalmixEM(galaxies, k = 2, maxit = 1000, epsilon = 1e-6)

# 3-component mixture
mix3 <- normalmixEM(galaxies, k = 3, maxit = 1000, epsilon = 1e-6)
```

### Plot: 2-Component Fit

```{r galaxies-mix2-plot}
hist(galaxies, breaks = 30, col = "grey85", border = "white",
     probability = TRUE, ylim = c(0, 0.00020),
     main = "2-Component Normal Mixture",
     xlab = "Velocity (km/s)")

xseq <- seq(min(galaxies) - 2000, max(galaxies) + 2000, length = 500)
for (j in 1:2) {
  lines(xseq, mix2$lambda[j] * dnorm(xseq, mix2$mu[j], mix2$sigma[j]),
        col = j + 1, lwd = 2)
}
# Total mixture density
total2 <- mix2$lambda[1] * dnorm(xseq, mix2$mu[1], mix2$sigma[1]) +
          mix2$lambda[2] * dnorm(xseq, mix2$mu[2], mix2$sigma[2])
lines(xseq, total2, col = 1, lwd = 2, lty = 2)
legend("topright", c("Comp 1", "Comp 2", "Total"),
       col = c(2, 3, 1), lwd = 2, lty = c(1, 1, 2), cex = 0.8)
```

### Plot: 3-Component Fit

```{r galaxies-mix3-plot}
hist(galaxies, breaks = 30, col = "grey85", border = "white",
     probability = TRUE, ylim = c(0, 0.00020),
     main = "3-Component Normal Mixture",
     xlab = "Velocity (km/s)")

for (j in 1:3) {
  lines(xseq, mix3$lambda[j] * dnorm(xseq, mix3$mu[j], mix3$sigma[j]),
        col = j + 1, lwd = 2)
}
total3 <- Reduce("+", lapply(1:3, function(j)
  mix3$lambda[j] * dnorm(xseq, mix3$mu[j], mix3$sigma[j])))
lines(xseq, total3, col = 1, lwd = 2, lty = 2)
legend("topright", c("Comp 1", "Comp 2", "Comp 3", "Total"),
       col = c(2, 3, 4, 1), lwd = 2, lty = c(1, 1, 1, 2), cex = 0.8)
```

## AIC / BIC Comparison

For a $K$-component Normal mixture, we have $k = 3K - 1$ parameters.

```{r galaxies-aic-bic}
n_gal <- length(galaxies)

# Log-likelihood for a fitted mixture
loglik_mix <- function(fit) {
  ll <- 0
  for (i in 1:length(fit$x)) {
    li <- sum(fit$lambda * dnorm(fit$x[i], fit$mu, fit$sigma))
    ll <- ll + log(li)
  }
  ll
}

ll2 <- loglik_mix(mix2)
ll3 <- loglik_mix(mix3)

k2 <- 3 * 2 - 1   # = 5 parameters
k3 <- 3 * 3 - 1   # = 8 parameters

aic2 <- -2 * ll2 + 2 * k2
aic3 <- -2 * ll3 + 2 * k3
bic2 <- -2 * ll2 + k2 * log(n_gal)
bic3 <- -2 * ll3 + k3 * log(n_gal)

cat("2-Component:  AIC =", round(aic2, 2), " BIC =", round(bic2, 2), "\n")
cat("3-Component:  AIC =", round(aic3, 2), " BIC =", round(bic3, 2), "\n")
cat("\nPreferred by AIC:", ifelse(aic3 < aic2, "3-component", "2-component"), "\n")
cat("Preferred by BIC:", ifelse(bic3 < bic2, "3-component", "2-component"), "\n")
```

**Discussion:** The 3-component model should have lower AIC/BIC, agreeing with what the KDE suggested (multiple modes visible in the data).

\newpage

# Mixture Models — Simulated Data

## Generate the Mixture Sample

```{r sim-mixture}
set.seed(1138)
n <- 100
lambda_true <- 0.3

component <- rbinom(n, size = 1, prob = lambda_true)
mixture_sample <- ifelse(component == 1,
                         rnorm(n, mean = 10, sd = 5),
                         rnorm(n, mean = 30, sd = 10))

hist(mixture_sample, breaks = 20, col = "lightblue",
     probability = TRUE, main = "Simulated 2-Component Mixture",
     xlab = "x")
```

**True parameters:** $\lambda = 0.3$, $\mu_1 = 10$, $\sigma_1 = 5$, $\mu_2 = 30$, $\sigma_2 = 10$.

## EM Algorithm — Manual Implementation

Here is the EM code from lectures. Let's walk through what every line does:

```{r em-manual}
twoMix <- function(x) {
  # ── Initialise ──
  lambda <- 0.5                   # start with equal mixing
  mu     <- range(x)              # mu1 = min(x), mu2 = max(x)
  sigma  <- rep(sd(x), 2)        # both sigmas = overall SD
  
  repeat {
    oldp <- c(lambda, mu, sigma)  # store old parameters
    
    # ── E-step: compute responsibilities ──
    # gamma_i = P(point i belongs to component 1)
    gamma <- lambda * dnorm(x, mu[1], sigma[1])
    gamma <- gamma / (gamma + (1 - lambda) * dnorm(x, mu[2], sigma[2]))
    
    # ── M-step: update parameters ──
    mu[1]    <- sum(gamma * x) / sum(gamma)
    mu[2]    <- sum((1 - gamma) * x) / sum(1 - gamma)
    sigma[1] <- sqrt(sum(gamma * (x - mu[1])^2) / sum(gamma))
    sigma[2] <- sqrt(sum((1 - gamma) * (x - mu[2])^2) / sum(1 - gamma))
    lambda   <- sum(gamma) / length(x)
    
    # ── Check convergence ──
    diff <- sum(abs(c(lambda, mu, sigma) - oldp))
    if (diff < 0.001) break
  }
  
  return(list(lambda = lambda, mu = mu, sigma = sigma))
}

result <- twoMix(mixture_sample)
cat("Estimated lambda:", result$lambda, "\n")
cat("Estimated mu:    ", result$mu, "\n")
cat("Estimated sigma: ", result$sigma, "\n")
cat("\nTrue values: lambda=0.3, mu=(10,30), sigma=(5,10)\n")
```

\examtip{The EM algorithm is sensitive to starting values. The code initialises $\mu$ at the min and max of the data, which works well when components are separated. For badly separated components, try multiple random starts.}

## How Initial Values Affect the Outcome

```{r em-initial-values}
# Try different starting configurations
twoMix_custom <- function(x, lambda_init, mu_init, sigma_init) {
  lambda <- lambda_init
  mu     <- mu_init
  sigma  <- sigma_init
  iter   <- 0
  
  repeat {
    oldp <- c(lambda, mu, sigma)
    iter <- iter + 1
    
    gamma <- lambda * dnorm(x, mu[1], sigma[1])
    gamma <- gamma / (gamma + (1 - lambda) * dnorm(x, mu[2], sigma[2]))
    
    mu[1]    <- sum(gamma * x) / sum(gamma)
    mu[2]    <- sum((1 - gamma) * x) / sum(1 - gamma)
    sigma[1] <- sqrt(sum(gamma * (x - mu[1])^2) / sum(gamma))
    sigma[2] <- sqrt(sum((1 - gamma) * (x - mu[2])^2) / sum(1 - gamma))
    lambda   <- sum(gamma) / length(x)
    
    diff <- sum(abs(c(lambda, mu, sigma) - oldp))
    if (diff < 0.001 || iter > 500) break
  }
  
  list(lambda = lambda, mu = mu, sigma = sigma, iterations = iter)
}

# Good starting values
r1 <- twoMix_custom(mixture_sample, 0.5, c(5, 35), c(8, 8))
cat("Start 1 (good):  mu =", round(r1$mu, 2),
    " sigma =", round(r1$sigma, 2),
    " lambda =", round(r1$lambda, 3), "\n")

# Mediocre starting values
r2 <- twoMix_custom(mixture_sample, 0.5, c(20, 25), c(15, 15))
cat("Start 2 (close): mu =", round(r2$mu, 2),
    " sigma =", round(r2$sigma, 2),
    " lambda =", round(r2$lambda, 3), "\n")

# Identical starting means (degenerate)
r3 <- twoMix_custom(mixture_sample, 0.5, c(20, 20), c(10, 10))
cat("Start 3 (same):  mu =", round(r3$mu, 2),
    " sigma =", round(r3$sigma, 2),
    " lambda =", round(r3$lambda, 3), "\n")
```

**Key insight:** When both starting means are the same, the algorithm can't distinguish the components and may converge to a degenerate solution (one component swallows everything). Always use **different** starting means.

## Watch the EM Converge — Animated Iterations

```{r em-animated, fig.height=8}
# EM with plotting at each iteration
x <- mixture_sample
lambda <- 0.5
mu     <- range(x)
sigma  <- rep(sd(x), 2)

par(mfrow = c(3, 3), mar = c(3, 3, 2, 1))
xseq <- seq(min(x) - 10, max(x) + 10, length = 300)

for (step in 1:9) {
  # E-step
  gamma <- lambda * dnorm(x, mu[1], sigma[1])
  gamma <- gamma / (gamma + (1 - lambda) * dnorm(x, mu[2], sigma[2]))
  
  # M-step
  mu[1]    <- sum(gamma * x) / sum(gamma)
  mu[2]    <- sum((1 - gamma) * x) / sum(1 - gamma)
  sigma[1] <- sqrt(sum(gamma * (x - mu[1])^2) / sum(gamma))
  sigma[2] <- sqrt(sum((1 - gamma) * (x - mu[2])^2) / sum(1 - gamma))
  lambda   <- sum(gamma) / length(x)
  
  # Plot
  hist(x, breaks = 20, probability = TRUE, col = "grey90", border = "white",
       main = paste("Iteration", step), xlab = "", cex.main = 0.9)
  curve(lambda * dnorm(x, mu[1], sigma[1]), add = TRUE, col = 2, lwd = 2)
  curve((1 - lambda) * dnorm(x, mu[2], sigma[2]), add = TRUE, col = 3, lwd = 2)
  curve(lambda * dnorm(x, mu[1], sigma[1]) +
        (1 - lambda) * dnorm(x, mu[2], sigma[2]),
        add = TRUE, col = 1, lwd = 2, lty = 2)
}
```

You can see the two component densities (red and green) gradually shift and reshape until they settle on the final fit.

## MLE Approach

We can also fit the mixture using direct maximum likelihood with `mle()`:

```{r mle-mixture}
library(stats4)

negloglik2 <- function(lambda, mu1, s1, mu2, s2) {
  -sum(log(lambda * dnorm(mixture_sample, mu1, s1) +
           (1 - lambda) * dnorm(mixture_sample, mu2, s2)))
}

mle_fit <- mle(negloglik2,
               start = list(lambda = 0.5, mu1 = 10, s1 = 5, mu2 = 30, s2 = 10),
               method = "L-BFGS-B",
               lower = c(0.01, -Inf, 0.1, -Inf, 0.1),
               upper = c(0.99, Inf, Inf, Inf, Inf))

cat("MLE estimates:\n")
print(coef(mle_fit))
```

\newpage

# Turtle Data — Circular Mixture

## Load the Turtle Data

```{r turtle-data}
library(CircNNTSR)
data(Turtles)
head(Turtles)

# The data is in degrees or radians — check:
turtle_vals <- Turtles$theta  # adjust column name if needed
hist(turtle_vals, breaks = 20, col = "lightgreen",
     main = "Turtle Swimming Directions", xlab = "Direction")
```

## Fit with MLE

```{r turtle-mle, eval=FALSE}
# Using mle() — may need careful starting values
negloglik_turtle <- function(lambda, mu1, s1, mu2, s2) {
  -sum(log(lambda * dnorm(turtle_vals, mu1, s1) +
           (1 - lambda) * dnorm(turtle_vals, mu2, s2)))
}

mle_turtle <- mle(negloglik_turtle,
                  start = list(lambda = 0.5, mu1 = 50, s1 = 30,
                               mu2 = 200, s2 = 30),
                  method = "L-BFGS-B",
                  lower = c(0.01, -Inf, 0.1, -Inf, 0.1),
                  upper = c(0.99, Inf, Inf, Inf, Inf))
print(coef(mle_turtle))
```

## Fit with Manual EM

```{r turtle-em}
turtle_em <- twoMix(turtle_vals)
cat("EM results for Turtles:\n")
cat("lambda:", turtle_em$lambda, "\n")
cat("mu:    ", turtle_em$mu, "\n")
cat("sigma: ", turtle_em$sigma, "\n")
```

## Fit with normalmixEM (Gold Standard)

```{r turtle-normalmixem}
set.seed(42)
turtle_mix <- normalmixEM(turtle_vals, k = 2, maxit = 1000)

cat("\nnormalmixEM results:\n")
cat("lambda:", turtle_mix$lambda, "\n")
cat("mu:    ", turtle_mix$mu, "\n")
cat("sigma: ", turtle_mix$sigma, "\n")

# Compare all three methods
cat("\n--- Comparison ---\n")
cat("Manual EM:     mu =", round(turtle_em$mu, 2),
    "sigma =", round(turtle_em$sigma, 2), "\n")
cat("normalmixEM:   mu =", round(turtle_mix$mu, 2),
    "sigma =", round(turtle_mix$sigma, 2), "\n")
```

**They should agree** (up to label switching — component 1 and 2 might be swapped).

\examtip{Label switching: the EM algorithm doesn't care which component is ``1'' and which is ``2''. If your manual EM gives $\mu_1=200, \mu_2=50$ and normalmixEM gives $\mu_1=50, \mu_2=200$, they're the same solution — just relabelled.}

\newpage

# ThreeMix Data — Fitting 2, 3, and 4 Components

## Load and Explore

```{r threemix-data}
# Load the ThreeMix data
threemix <- read.csv("ThreeMix.csv")
x_three <- threemix$x

hist(x_three, breaks = 30, col = "grey85", border = "white",
     probability = TRUE,
     main = "ThreeMix Data", xlab = "x")
lines(density(x_three), col = "blue", lwd = 2)
```

## Fit 2, 3, and 4 Component Mixtures

```{r threemix-fit}
set.seed(123)
mix_2 <- normalmixEM(x_three, k = 2, maxit = 2000, epsilon = 1e-6)
mix_3 <- normalmixEM(x_three, k = 3, maxit = 2000, epsilon = 1e-6)
mix_4 <- normalmixEM(x_three, k = 4, maxit = 2000, epsilon = 1e-6)
```

## Plot Each Fit

```{r threemix-plots, fig.height=9}
par(mfrow = c(3, 1), mar = c(4, 4, 2, 1))
xseq <- seq(min(x_three) - 5, max(x_three) + 5, length = 500)

for (info in list(list(mix_2, 2), list(mix_3, 3), list(mix_4, 4))) {
  fit <- info[[1]]; K <- info[[2]]
  
  hist(x_three, breaks = 30, col = "grey90", border = "white",
       probability = TRUE,
       main = paste0(K, "-Component Mixture"), xlab = "x")
  
  total <- rep(0, length(xseq))
  for (j in 1:K) {
    comp <- fit$lambda[j] * dnorm(xseq, fit$mu[j], fit$sigma[j])
    lines(xseq, comp, col = j + 1, lwd = 2)
    total <- total + comp
  }
  lines(xseq, total, col = 1, lwd = 2, lty = 2)
}
```

## AIC and BIC Comparison

```{r threemix-aic-bic}
n_three <- length(x_three)

calc_loglik <- function(fit) {
  ll <- 0
  for (i in seq_along(fit$x)) {
    li <- sum(fit$lambda * dnorm(fit$x[i], fit$mu, fit$sigma))
    ll <- ll + log(li)
  }
  ll
}

results <- data.frame(
  K          = c(2, 3, 4),
  Parameters = c(3*2-1, 3*3-1, 3*4-1),
  LogLik     = c(calc_loglik(mix_2), calc_loglik(mix_3), calc_loglik(mix_4))
)
results$AIC <- -2 * results$LogLik + 2 * results$Parameters
results$BIC <- -2 * results$LogLik + results$Parameters * log(n_three)

print(results)

cat("\nBest by AIC:", results$K[which.min(results$AIC)], "components\n")
cat("Best by BIC:", results$K[which.min(results$BIC)], "components\n")
```

\examtip{For the AIC/BIC calculation, a $K$-component Normal mixture has $k = 3K - 1$ free parameters: $K$ means + $K$ standard deviations + $(K-1)$ mixing proportions (they must sum to 1, so only $K-1$ are free).}

\newpage

# Precipitation Data — Built-in

## Explore

```{r precip-data}
data(precip)
cat("Number of cities:", length(precip), "\n")
summary(precip)

hist(precip, breaks = 20, col = "lightblue", border = "white",
     probability = TRUE, main = "Annual Rainfall — 70 US Cities",
     xlab = "Precipitation (inches)")
lines(density(precip), col = "blue", lwd = 2)
```

## Fit 1, 2, and 3 Component Models

```{r precip-fit}
set.seed(42)

# 1-component = just a Normal distribution
precip_1_mu    <- mean(precip)
precip_1_sigma <- sd(precip)
ll_1 <- sum(dnorm(precip, precip_1_mu, precip_1_sigma, log = TRUE))

# 2-component mixture
precip_mix2 <- normalmixEM(precip, k = 2, maxit = 2000)

# 3-component mixture
precip_mix3 <- normalmixEM(precip, k = 3, maxit = 2000)
```

```{r precip-plot, fig.height=7}
par(mfrow = c(2, 1), mar = c(4, 4, 2, 1))
xseq <- seq(0, 80, length = 500)

# 2-component plot
hist(precip, breaks = 20, col = "grey90", border = "white",
     probability = TRUE, main = "2-Component Mixture", xlab = "Precipitation")
total2 <- rep(0, length(xseq))
for (j in 1:2) {
  comp <- precip_mix2$lambda[j] * dnorm(xseq, precip_mix2$mu[j], precip_mix2$sigma[j])
  lines(xseq, comp, col = j + 1, lwd = 2)
  total2 <- total2 + comp
}
lines(xseq, total2, lty = 2, lwd = 2)

# 3-component plot
hist(precip, breaks = 20, col = "grey90", border = "white",
     probability = TRUE, main = "3-Component Mixture", xlab = "Precipitation")
total3 <- rep(0, length(xseq))
for (j in 1:3) {
  comp <- precip_mix3$lambda[j] * dnorm(xseq, precip_mix3$mu[j], precip_mix3$sigma[j])
  lines(xseq, comp, col = j + 1, lwd = 2)
  total3 <- total3 + comp
}
lines(xseq, total3, lty = 2, lwd = 2)
```

## AIC / BIC — Which Model is Best?

```{r precip-compare}
n_precip <- length(precip)

ll_2 <- calc_loglik(precip_mix2)
ll_3 <- calc_loglik(precip_mix3)

precip_results <- data.frame(
  K          = 1:3,
  Parameters = c(2, 5, 8),
  LogLik     = c(ll_1, ll_2, ll_3)
)
precip_results$AIC <- -2 * precip_results$LogLik + 2 * precip_results$Parameters
precip_results$BIC <- -2 * precip_results$LogLik +
                      precip_results$Parameters * log(n_precip)

print(precip_results)

cat("\nBest by AIC:", precip_results$K[which.min(precip_results$AIC)], "component(s)\n")
cat("Best by BIC:", precip_results$K[which.min(precip_results$BIC)], "component(s)\n")
```

**Discussion:** BIC often favours fewer components than AIC. If AIC says 3 but BIC says 2, a 2-component model is a reasonable, parsimonious choice.

\newpage

# Complete Exam Summary

## Key Concepts at a Glance

\begin{center}
\begin{tabular}{|l|p{10cm}|}
\hline
\textbf{Concept} & \textbf{What to Remember} \\
\hline
KDE & Smooth histogram. Bandwidth $h$ controls smoothness. Use \texttt{density()} in R. \\
\hline
Bandwidth & \texttt{nrd0} = rule of thumb (assumes unimodal). \texttt{ucv} = cross-validation (better for multimodal). \\
\hline
Bayes Classifier & $P(Y{=}1|x) = \frac{\pi_1 f_1(x)}{\pi_0 f_0(x) + \pi_1 f_1(x)}$. Classify as 1 if $> 0.5$. \\
\hline
Mixture Model & $f(x) = \sum_k \lambda_k \phi(x; \mu_k, \sigma_k)$. Components are sub-populations. \\
\hline
EM Algorithm & E-step: compute responsibilities. M-step: update parameters. Repeat. Sensitive to initial values. \\
\hline
AIC & $-2\log L + 2k$. Lower = better. \\
\hline
BIC & $-2\log L + k\log n$. Lower = better. Penalises more than AIC. \\
\hline
Parameters & $K$-component Normal mixture: $k = 3K - 1$ free parameters. \\
\hline
\end{tabular}
\end{center}

## Common Exam Mistakes to Avoid

1. **Forgetting `probability = TRUE`** in `hist()` when overlaying densities — without it, the y-axis is counts, not density, and your curves won't match.
2. **Using `nrd0` for multimodal data** — it oversmooths. Mention `ucv` or `SJ` instead.
3. **Confusing the number of free parameters** — for $K$ components it's $3K - 1$ (not $3K$), because the $\lambda$'s must sum to 1.
4. **Ignoring label switching** — if your EM gives components in a different order than `normalmixEM`, they're still the same solution.
5. **Not trying multiple initial values** — EM can get stuck in local maxima. Always mention this in an exam answer.
